Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-18 _tex/paper2.tex
Built site for gh-pages

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  1p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\setstretch{2}
\section{Introduction}\label{sec-introduction}

@@

Figure~\ref{fig-discriminal_process} illustrates hypothetical
discriminal processes along a quality trait continuum for two written
texts. The figure indicates that the modal discriminal process for Text
B is positioned further along the continuum than that of Text A
\((S_{B} > S_{A})\), suggesting that Text B exhibits higher quality.
Additionally, the figure highlights that Text B has a broader
distribution compared to Text A, which arises from its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the hypothetical texts depicted in panel
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion rests on two key observations: the positive
difference between their modal discriminal processes
\((S_{B} - S_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over Text A.

@@

Thurstone observed that the general form of the theory, outlined in
Section~\ref{sec-thurstone_theory}, created a trait scaling problem. The
model required estimating more ``unknown'' parameters than the available
pairwise comparisons \citep[pp.~267]{Thurstone_1927b}. To address this
issue and facilitate the practical application of the theory, he
developed five cases derived from this general form, each case
progressively incorporating additional simplifying assumptions into the
model.

In Case I, Thurstone assumed that pairs of stimuli maintained a constant
correlation across all comparisons. In Case II, he allowed multiple
judges to make comparisons instead of restricting evaluations to a
single judge. In Case III, he introduced the assumption of zero
correlation between stimuli. In Case IV, he assumed stimuli exhibited
similar dispersions. Finally, in Case V, he replaced this assumption
with the condition that stimuli had equal discriminal dispersions.
Table~\ref{tbl-thurstone_cases} summarizes the assumptions of the
general form and the five cases. For an in-depth discussion of these
cases and their progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248--253]{Bramley_2008}.

@@

Despite relying on the most extensive set of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one key difference: while Case V assumes a
Normal distribution for the stimuli's discriminal processes, the BTL
model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions share similar statistical properties, differing only by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I}.

However, Thurstone originally developed Case V to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
untested application, stating that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
that some assumptions could prove problematic when researchers assess
complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve such traits and stimuli, two main
assumptions of Case V and, by extension, of the BTL model may not
consistently hold in theory or practice: the assumption of equal
dispersion and zero correlation between stimuli.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, exerting a
direct influence on the outcome of pairwise comparisons.
Figure~\ref{fig-dispersion} illustrates this idea using a thought
experiment in which a researcher can observe the discriminal processes
for the texts shown in Figure~\ref{fig-discriminal_process}.
Furthermore, the figure assumes that the discriminal dispersion for Text
A remains constant and that the texts are uncorrelated \((\rho=0)\). The
figure reveals that an increase in the uncertainty associated with the
perception of Text B in comparison to Text A,
\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B > A)\), ultimately influencing the comparison
outcome. Additionally, the figure reveals that when the discriminal
dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
discriminal difference is more likely to favor Text B over Text A
(shaded gray area), compared to situations where their dispersions
differ.

In experimental practice, however, this process occurs in reverse.
Researchers first observe the comparison outcome and then use the BTL
model to infer the discriminal difference between the stimuli and their
respective discriminal processes \citep[pp.~373]{Thurstone_1927a}. For
instance, when researchers observe a large sample of outcomes favoring
Text B over Text A and correctly assume equal dispersions between the
texts, the BTL model estimates a discriminal difference distribution
that accurately represents the ``true'' discriminal difference of the
texts. This scenario is illustrated in Figure~\ref{fig-dispersion},
where the model's discriminal difference distribution aligns with the
``true'' distribution, represented by the thick continuous line
corresponding to \(\sigma_{B}-\sigma_{A}=0\). The accuracy of this
estimated discriminal difference ensures reliable estimates of the
texts' discriminal processes {[}(citation
needed?)\{style=``color:red;''\}. Therefore, it is reasonable to assume
that the outcome's ability to reflect the ``true'' differences between
stimuli largely depends on the validity of the model's assumptions,
particularly the assumption of equal dispersions.

However, Thurstone contended that the assumption of equal dispersions
may not hold when researchers assess complex traits or heterogeneous
stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli
can introduce judgment discrepancies due to their unique features
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, evidence of this violation may already exist in the CJ
literature as misfit statistics, which measure judgment discrepancies
associated with specific stimuli
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For example, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than those
involving other texts
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
These discrepancies, in turn, suggest that the discriminal differences
for ``misfit'' texts have broader distributions, indicating that their
discriminal processes may also exhibit more variation than that of other
texts. A similar reasoning applies to ``misfit'' judges, whose
evaluations deviate substantially from the shared consensus due to the
unique characteristics of the stimuli or the judges themselves.
Moreover, these ``misfit'' judges and their deviations can introduce
additional statistical and measurement issues, which we discuss in
Section~\ref{sec-theory-issue1b}.

Therefore, incorrectly assuming equal dispersions between stimuli is not
without harm, as it can cause the BTL model to introduce significant
statistical and measurement issues. For instance, the model may
overestimate how accurately the outcome reflects the ``true''
discriminal differences between stimuli. This overestimation can lead
researchers to draw spurious conclusions about these differences
\citep[pp.~370]{McElreath_2020} and, by extension, of the stimuli
underlying discriminal processes. Figure~\ref{fig-dispersion}
illustrates this issue when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
follows a discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).
Moreover, if researchers recognize that misfit statistics highlight
critical differences in dispersions, the usual CJ practice of excluding
stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
can unintentionally discard valuable information. This exclusion can
introduce bias into trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases are often unpredictable, as they depend on
which stimuli are excluded from the analysis.

@@

Denoted by \(\rho\), the correlation measures the dependence of a
judge's perception of the trait in one stimulus on his perception of the
same trait in another. Like the discriminal dispersions, this
correlation shapes the distribution of the discriminal difference and
directly influences the outcomes of pairwise comparisons.
Figure~\ref{fig-correlation} illustrates this concept, again assuming
the researcher can observe the discriminal processes for the texts shown
in Figure~\ref{fig-discriminal_process}. The figure also considers that
the discriminal dispersions for both texts remain constant.

Figure~\ref{fig-correlation} reveals that as the correlation between the
texts increases, the distribution of their discriminal difference
becomes narrower. This narrowing affects the area under the curve where
the discriminal difference distinctly favors Text B over Text A, denoted
as \(P(B > A)\), thus influencing the comparison outcome. Furthermore,
the figure shows that when two texts are independent or uncorrelated
\((\rho=0)\), their discriminal difference is less likely to favor Text
B over Text A (shaded gray area), compared to scenarios when the texts
are highly correlated.

However, in experimental practice, researchers typically follow this
process in reverse. For example, when they observe a large sample of
outcomes favoring Text B over Text A and correctly assume zero
correlation between the texts, the BTL model estimates a discriminal
difference distribution that accurately represents the ``true''
discriminal difference of the texts. This scenario is illustrated with
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
The accuracy of this discriminal difference estimation, in turn, ensures
reliable estimates for the discriminal process of the texts {[}(citation
needed?)\{style=``color:red;''\}.

Notably, Thurstone's Case V and the BTL model assume independent
discriminal processes across comparisons. Thurstone attributed this
independence to the cancellation of potential judges' biases, driven by
two opposing and equally weighted effects occurring during the pairwise
comparisons \citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978}
mathematically demonstrated this cancellation using the BTL model under
the assumption of discriminal processes with additive biases. However,
it is easy to imagine at least two scenarios where the zero correlation
assumption almost certainly does not hold: when the pairwise comparison
involves multidimensional, complex traits with heterogeneous stimuli and
when an additional hierarchical structure is relevant to the stimuli.

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluate these traits, they often rely
on various intricate characteristics of the stimuli to form their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, which are unlikely to be equally
weighted or opposing, can unevenly influence judges' perceptions,
creating biases in their judgments and, ultimately, introducing
dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
more weight on its grammatical accuracy than other judges, ultimately
favoring texts with fewer errors but weaker arguments. While direct
evidence for this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the idea that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the presence of such hierarchical structures, the
statistical handling of this extra source of dependency between stimuli
has been inadequate. For example, when CJ data includes multiple samples
of stimuli from the same individuals, researchers often rely on
(average) estimated BTL scores to conduct subsequent analyses and tests
at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in Section~\ref{sec-theory-issue2}.

{In any case, similar to Section~\ref{sec-theory-issue1a}, incorrectly
assuming zero correlation between stimuli can lead the BTL model to
introduce various statistical and measurement issues. For instance, the
model could over- or underestimate the accuracy of the outcome in
reflecting the ``true'' discriminal differences between stimuli. This
over- underestimation may result in spurious inferences about these
differences and, by extension, about the stimuli's discriminal processes
\citep[pp.~341]{Hoyle_et_al_2023}. Figure~\ref{fig-correlation} also
illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for \(\rho=0\), while
the ``true'' discriminal difference follows any discontinuous line where
\(\rho \neq 0\). This missaligment can be due to the overlook of
additional relevant traits, such as judges' biases, which cause
dimensional mismatches in the BTL model, artificially inflating the
reliability of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, even
worse, introduce bias into the trait's estimates \citep{Ackerman_1989}.
Furthermore, researchers who exclude judges based on misfit statistics
can risk discarding valuable information, further biasing the trait's
estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Lastly,
researchers who fail to account for hierarchical (grouping) structures
can reduce the precision of model parameter estimates, which may amplify
the overestimation of the trait's reliability
\citep[pp.~482]{Hoyle_et_al_2023}.}

@@

{In aggregate, researchers' inadequate handling of violations to the
assumptions of equal dispersion and zero correlation between stimuli,
along with the apparent disconnect between CJ's approach to trait
measurement and hypothesis testing, can undermine the reliability of the
trait estimates and ultimately compromise its validity
\citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
systematic and integrated approach to examining what happens when judges
compare two stimuli could offer several statistical and measurement
benefits, including addressing these issues.}

@@

This section presents a theoretical model for CJ that extends
Thurstone's theory. The model systematically incorporates all factors
involved when judges make pairwise comparisons. Additionally, the
section develops the statistical translation of the theoretical model
based on assumptions informed by the CJ theory.

@@

@@

\includegraphics[width=0.29\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A1.png}

@@

\caption{\label{fig-CJ_TM_A1}Theoretical model A1\$}

@@

@@

\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A2.png}

@@

\caption{\label{fig-CJ_TM_A2}Theoretical model A2\$}

@@

@@

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A3.png}

@@

\caption{\label{fig-CJ_TM_A3}Theoretical model A3\$}

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  review,
  1p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\section{Introduction}\label{sec-introduction}

@@

Figure~\ref{fig-discriminal_process} illustrates hypothetical
discriminal processes along a quality trait continuum for two written
texts. The figure indicates that the modal discriminal process for Text
B is positioned further along the continuum than that of Text A
\((T_{B} > T_{A})\), suggesting that Text B exhibits higher quality.
Additionally, the figure highlights that Text B has a broader
distribution compared to Text A, which arises from its larger
discriminal dispersion \((\sigma_{B} > \sigma_{A})\).

@@

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the hypothetical texts depicted in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion rests on two key observations: the positive
difference between their modal discriminal processes
\((T_{B} - T_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over Text A.

@@

Thurstone noted that the general form of the theory, as delineated in
Section~\ref{sec-thurstone_theory}, gave rise to a trait scaling
problem. The model required estimating more ``unknown'' parameters than
the available pairwise comparisons \citep[pp.~267]{Thurstone_1927b}. To
address this issue and facilitate the practical implementation of the
theory, he developed five cases derived from this general form, each
case progressively incorporated additional simplifying assumptions into
the model.

In Case I, Thurstone postulated that pairs of stimuli would maintain a
constant correlation across all comparisons. In Case II, he allowed
multiple judges to undertake comparisons instead of confining
evaluations to a single judge. In Case III, he posited that there was no
correlation between stimuli. In Case IV, he assumed that the stimuli
exhibited similar dispersions. Finally, in Case V, he replaced this
assumption with the condition that stimuli had equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes the assumptions
of the general form and the five cases. For a detailed discussion of
these cases and their progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248--253]{Bramley_2008}.

@@

Despite relying on the most extensive set of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one notable distinction: whereas Case V
assumes a Normal distribution for the stimuli's discriminal processes,
the BTL model uses the more mathematically tractable Logistic
distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions exhibit analogous statistical properties, differing only
by a scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I}.

However, Thurstone originally developed Case V to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
untested application, stating that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}. Furthermore,
he acknowledged that some assumptions could prove problematic when
researchers assess complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve such traits and stimuli, two main
assumptions of Case V and, by extension, of the BTL model may not
consistently hold in theory or practice, namely the assumption of equal
dispersion and zero correlation between stimuli.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, exerting a
direct influence on the outcome of pairwise comparisons.
Figure~\ref{fig-dispersion} presents a thought experiment to illustrate
this idea. In this experiment, a researcher can observe the discriminal
processes for the texts depicted in
Figure~\ref{fig-discriminal_process}. Furthermore, the figure assumes
that the discriminal dispersion for Text A remains constant and that the
texts are uncorrelated \((\rho=0)\). The figure reveals that an increase
in the uncertainty associated with the perception of Text B in
comparison to Text A, \((\sigma_{B}-\sigma_{A})\), broadens the
distribution of their discriminal difference. This broadening affects
the probability area where the discriminal difference distinctly favors
Text B over Text A, expressed as \(P(B > A)\), ultimately influencing
the comparison outcome. Additionally, the figure reveals that when the
discriminal dispersions of the texts are equal
\((\sigma_{B}-\sigma_{A}=0)\), the discriminal difference is more likely
to favor Text B over Text A (shaded gray area), compared to situations
where their dispersions differ.

In experimental practice, however, this process occurs in reverse.
Researchers first observe the comparison outcome and then use the BTL
model to infer the discriminal difference between the stimuli and their
respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
Therefore, it is reasonable to assume that the outcome's ability to
reflect the ``true'' differences between stimuli largely depends on the
validity of the model's assumptions, particularly the assumption of
equal dispersions. For instance, when researchers observe a sample of
outcomes favoring Text B over Text A and correctly assume equal
dispersions between the texts, the BTL model estimates a discriminal
difference distribution that accurately represents the ``true''
discriminal difference of the texts. This scenario is illustrated in
Figure~\ref{fig-dispersion}, where the model's discriminal difference
distribution aligns with the ``true'' distribution, represented by the
thick continuous line corresponding to \(\sigma_{B}-\sigma_{A}=0\). The
accuracy of these discriminal difference ensures reliable estimates for
the texts' discriminal processes {(citation needed?)}.

However, Thurstone argued that the assumption of equal dispersions may
not be applicable when researchers assess complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
and stimuli can introduce judgment discrepancies due to their unique
features
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, evidence of this violation may already be present in the CJ
literature in the form of misfit statistics, which measure judgment
discrepancies associated with specific stimuli
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For example, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than those
involving other texts
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
These discrepancies, in turn, suggest that the discriminal differences
for ``misfit'' texts have broader distributions, indicating that their
discriminal processes may also exhibit more variation than that of other
texts. A similar line of reasoning applies to the concept of ``misfit''
judges, whose evaluations deviate substantially from the shared
consensus due to the unique characteristics of the stimuli or the judges
themselves. These ``misfit'' judges and their associated deviations can
give rise to additional statistical and measurement issues, which we
discuss in more detail in Section~\ref{sec-theory-issue1b}.

Nevertheless, model misspecification, in the form of an erroneous
assumption of equal dispersions between stimuli, can give rise to
significant statistical and measurement issues. For instance, the model
may overestimate the degree to which the outcome accurately reflects the
``true'' discriminal differences between stimuli. This overestimation
can result in researchers drawing spurious conclusions about these
differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
illustrates this issue when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
follows any discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).

Additionally, if researchers recognize that misfit statistics highlight
these critical differences in dispersions, the conventional CJ practice
of excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
can unintentionally discard valuable information. Such exclusions can
introduce bias into trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases are often unpredictable, as they depend on
which stimuli are excluded from the analysis.

@@

The correlation, represented by the symbol \(\rho\), measures how much a
judge's perception of a specific trait in one stimulus depends on their
perception of the same trait in another. As with the discriminal
dispersions, this correlation shapes the distribution of the discriminal
difference, directly impacting the outcomes of pairwise comparisons.
Figure~\ref{fig-correlation} presents a similar thought experiment as in
Section~\ref{sec-theory-issue1a} to illustrate this idea. The
illustration now assumes that the discriminal dispersions for both texts
remain constant. The figure shows that as the correlation between the
texts increases, the distribution of their discriminal difference
becomes narrower. This narrowing affects the area under the curve where
the discriminal difference distinctly favors Text B over Text A, denoted
as \(P(B > A)\), thus influencing the comparison outcome. Furthermore,
the figure shows that when two texts are independent or uncorrelated
\((\rho=0)\), their discriminal difference is less likely to favor Text
B over Text A (shaded gray area) compared to scenarios where the texts
are highly correlated.

However, in practice, researchers approach this process in reverse. They
begin by observing the sample of outcomes favoring Text B over Text A
and then use the BTL model to estimate the discriminal difference and
the discriminal processes of the stimuli. Given that the BTL model
assumes independent discriminal processes across comparisons, if this
assumption holds, then the model estimates a discriminal difference
distribution that accurately reflects the ``true'' discriminal
difference of the texts. This scenario is illustrated with
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
Once more, the estimation accuracy of the discriminal difference ensures
reliable estimates for the discriminal processes of the texts {(citation
needed?)}.

Notably, Thurstone attributed the independence of stimuli to the
cancellation of potential judges' biases. He argued that this
cancellation resulted from two opposing and equally weighted effects
occurring during pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
\citet{Andrich_1978} provided a mathematical demonstration of this
cancellation using the BTL model under the assumption of discriminal
processes with additive biases. However, it is easy to imagine at least
two scenarios in which the zero correlation assumption is almost
certainly invalid: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli and when an
additional hierarchical structure is relevant to the stimuli.

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluate these traits, they often rely
on various intricate characteristics of the stimuli to form their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, which are unlikely to be equally
weighted or opposing, can exert an uneven influence on judges'
perceptions, creating biases in their judgments and, ultimately,
introducing dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
more weight on its grammatical accuracy than other judges, thereby
favoring texts with fewer errors but weaker arguments. While direct
evidence for this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the notion that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Despite the CJ literature
acknowledging the existence of such hierarchical structures, the
statistical handling of this additional source of dependence between
stimuli has been inadequate. For instance, when CJ data incorporates
multiple samples of stimuli from the same individuals, researchers
frequently rely on (average) estimated BTL scores to conduct subsequent
analyses and tests at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

In any case, similar to Section~\ref{sec-theory-issue1a}, model
misspecification due to an erroneous assumption of zero correlation
between stimuli can lead to significant statistical and measurement
issues. For instance, the model may over- or underestimate how
accurately the outcome reflects the ``true'' discriminal differences
between stimuli. Such inaccuracies can result in spurious inferences
about these differences and, by extension, about the stimuli's
discriminal processes. This scenario is illustrated in
Figure~\ref{fig-correlation}, where the model's discriminal difference
distribution aligns with the thick continuous line for \(\rho=0\), while
the ``true'' discriminal difference follows any discontinuous line where
\(\rho \neq 0\).

This misspecification may arise from neglecting additional relevant
traits, excluding judges based on misfit statistics, or ignoring
hierarchical (grouping) structures. Neglecting relevant traits, such as
judges' biases, can cause dimensional mismatches in the BTL model,
artificially inflating the trait's reliability
\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introducing bias into the
trait's estimates \citep{Ackerman_1989}. Excluding judges based on
misfit statistics risks discarding valuable information, which may
further bias the trait's estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Finally, ignoring
hierarchical structures may reduce the precision of model parameter
estimates, potentially amplifying the overestimation of the trait's
reliability \citep[pp.~482]{Hoyle_et_al_2023}.

@@

In aggregate, researchers' inadequate handling of violations to the
assumptions of equal dispersion and zero correlation between stimuli,
coupled with the apparent disconnect between CJ's approach to trait
measurement and hypothesis testing, can potentially compromise the
reliability of the trait estimates and, by extension, their validity
\citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
systematic and integrated approach to examine the factors influencing CJ
experiments could offer several statistical and measurement benefits,
including the ability to address these issues.

@@

This section introduces a theoretical model for Comparative Judgment
(CJ) that builds upon Thurstone's theory. The model systematically
accounts for the factors influencing judges during CJ experiments. These
theoretical and practical insights are then translated into a
statistical model.

@@

\paragraph{Figure 3}

@@

\includegraphics[width=0.2\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A1.png}

@@

\caption{\label{fig-CJ_TM_A1}}

@@

\paragraph{Figure 4}

@@

\includegraphics[width=0.3\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A2.png}

@@

\caption{\label{fig-CJ_TM_A2}}

@@

\paragraph{Figure 5}

@@

\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A3.png}

@@

\caption{\label{fig-CJ_TM_A3}}

\paragraph{Figure 6}

\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A4.png}

\caption{\label{fig-CJ_TM_A4}}

@@

\paragraph{Figure 7}

\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B1.png}
